Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-04-11 GWLDCI.tex
added pictures and responded to comments in sections 1-3

Often we take $V$ to be $[n]$, the cyclically ordered set of integers, $1 \ldots n$. In this case, we write $W = (\cP, [n])$. We introduce some notation to speak of vertices supporting a propagator, and the set of propagators supported on a vertex set.

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, [n])$.
\begin{enumerate}
\item For $p \in \cP$, let $V(p) = \{i_p, i_p+1, j_p, j_p+1\}$ be the set of vertices supporting $p$. Then, for $P \subseteq \cP$, the set $V(P) = \cup_{p \in P} V(p)$ is the vertex support of $P$.
\item For $V \subseteq [n]$, write $\Prop(V) = \{ p \in \cP | V(p) \cap V \neq \emptyset \} $.
\item For $P \subseteq \cP$, define $F(P) = V(P^c)^c$ to be the set of vertices in $[n]$ that do not support any propagators outside the set $P$.
\item The set of vertices that are not in the support of any propagotors is denoted  $F(\emptyset)$. Vertices in this set are called non-supporting.  
\end{enumerate}
\end{dfn}

@@

The first conditions states that there are at least four more vertices than propagators in an admissible Wilson Loop Diagram. The second imposes an upper bound on how densely the propagators can be fitted in the diagram. The third ensures that ensures that no propagators cross in the interior of the diagram. In other words, a Wilson loop diagram, $(\cP, [n])$ is admissible if and only if $n > \cP +4$, and has neither crossing propagators nor any pairs of propagators that start and end on the same pair of non-adjacent edges.

@@

\begin{dfn} \label{subdiagramdfn}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram. The weakly admissible diagram, $W'$ is a subdiagram of $W$, written $W' \subseteq W$, if \bas W' = (P, V); \quad P \subseteq \cP ; \quad V(P) \subseteq V \subseteq [n]\;.\eas
\end{dfn}

@@

\begin{dfn}\label{equivdfn} 
There is an equivalence relationship on the set of admissible Wilson loops diagrams given by the transititve closure the following binary relation: $W = (\cP, [n]) \sim W'= (\cP', n)$ if
\begin{enumerate}
\item There exist two different exact subdiagrams, $(P, V(P))$ and $(P', V(P'))$ of $W$ and $W'$ respectively such that $V(P) =  V(P')$.
\item The complementary subdiagrams are identical: $(\cP \setminus P, V(P)^c) = (\cP' \setminus P', V(P')^c)$.
\end{enumerate}
\end{dfn}

@@

Each Wilson loop diagram, $W = (\cP, [n])$ with $|\cP| = k$ is associated to a $k \times n$ matrix with non-zero real variable entries, called $C(W)$:

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any admissible Wilson loop diagram $W =(\cP, [n])$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ such that no subset supports fewer propagators than the vertices it contains.

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note that since vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$, $M(W)$ can also be thought of as the matroid realized by $C(W)$. \note{Can this note have a ref to your earlier paper?  If not then perhaps expand slightly as although I agree that it is true, I don't think it is quite so trivial.  Also, I think ``is also'' woulc be fine in place of ``can also be thought of as''}

@@

\begin{dfn}\label{WLDtriangulationdfn}
  Let $W = (\cP, [n])$ be an admissible Wilson loop diagram.  The \emph{polygon dissection} associated to $W$, denoted $\tau(W)$, is defined as follows.
  \begin{itemize}
  \item The vertices of $\tau(W)$ correspond to the edges of $W$.
  \item Labeling the vertices of $\tau(W)$ with the edge number of $W$gives a cyclic order to the vertices. Connecting consecutive vertices gives a graph theoretic cycle called the polygon of $\tau(W)$.
  \item Each propagator of $W$ defines a chord edge of $\tau(W)$; specifically,  a propagator $(i,j) \in \cP$ defines a chord connecting the vertices $i$ and $j$ in $\tau(W)$.
  \end{itemize}
\end{dfn}

@@

\begin{lem}\label{tausimpleplanarlem}
If $W = (\cP, [n])$ is an admissible Wilson loop diagram, then $\tau(W)$ is a simple planar graph whose outer face is a cycle. It is embedded such that the vertices all lie on this infinite face\footnote{That is, it is an \emph{outerplanar} graph}. These vertices are cyclically ordered, with a choice of first vertex giving it an additional compatible linear order.
\end{lem}

@@

\begin{proof}
We begin by giving an algorithm for the decomposition, then prove its uniqueness. Let $W = (\cP, [n])$, with $|\cP| = k$.

@@

Every triangulated piece $t$ of an admissible Wilson loop diagram $W$ corresponds to a subdiagram of $W$ by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $W_P = (P, V(P))$.  Conversely, given a subdiagram $W_P = (P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, as follows:
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ defined by cyclically consecutive elements of $V(P)$ as a subset of $[n]$
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$
  \item along with the outer edges of $\tau(W)$ for which both their end points are in the vertex set
  \end{itemize}
\end{itemize}
Note that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in $W_P$ itself.  In particular $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$.

@@

Conversely, suppose we have an exact subdiagram $(P, V(P))$ of $W = (\cP, [n])$ supported on $|V(P)| = |P|+3$ vertices, and let $t$ be the subgraph of $\tau(W)$ corresponding to $(P,V(P))$.

@@

\begin{cor} \label{uniqueproppartitioncor}
Any admissible Wilson loop diagram $W = (\cP, [n])$ can be uniquely decomposed into maximal exact subdiagrams. These maximal subdiagrams partition $\cP$.
\end{cor}

@@

\begin{dfn}\label{def prop flat} 
Let $W = (\cP,[n])$ be an admissible Wilson loop diagram and $P \subseteq \cP$. The set $F(P) := V(P^c)^c$ is called the {\em propagator flat} of $P$.
\end{dfn}
Thus in Lemma \ref{contractsubdiaglem} above we would contract $M(W)$ by the propagator flat $F(P^c)$ of the {\em complement} of $P$ in order to study $(P,V(P))$. The justification for this notation is given by (1) of the next lemma: the set $F(P)$ consists of vertices which {\em only} support propagators in $P$ (or no propagators at all).

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be a Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $F(P^c)$ is not an independent flat.
\item If $(P,V(P))$ is a maximal exact subdiagram in $W$, then $F(P^c)$ is a cyclic flat.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                          \item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate}
\end{lem}

@@

\begin{lem} \label{contractsubdiaglem} \cite[Theorem 3.33]{wilsonloop} 
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram and $P \subseteq \cP$. If the set $V(P)^c$ has rank $|P^c|$, then the matroid defined by the subdiagram $(P, V(P))$ is equal to the contraction $M(W)/V(P)^c$.
\end{lem}

@@

\begin{thm} \label{exactuniformthm}
Let $W':= (P, V(P))$ be a subdiagram of an admissible Wilson loop diagram $W= (\cP, [n])$. Then $W'$ is an exact subdiagram if and only if $M(W')$ is a uniform matroid of rank $|P|$.
\end{thm}

@@

\begin{dfn}
Given a Wilson loop diagram $W = (\cP, [n])$, define the positroid cell associated to a Wilson loop diagram, $\Sigma(W)$, to be the cell in the CW complex on $\Gr(|\cP|, n)$ defined by the positroid $M(W)$.
\end{dfn}

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, [n])$ and $W'= (\cP', [n])$ be two Wilson Loop diagrams. They define the same matroid if and only if $W \sim W'$.
\end{thm}

@@

Assume that $W$ and $W'$ are equivalent. Without loss of generality, write $W = (P \cup R, [n])$ and $W' = (P \cup R', [n])$, where $P \subset \cP \cap \cP'$ and $(R, V(R))$ and $(R', V(R'))$ are two maximally exact subdiagrams, with $R \neq R'$, but $V(R) = V(R')$. If this is not the case, one may always find a family of diagram, $\{W_i\}$ satisfying this condition and forming a transitive chain connecting $W$ to $W'$ in the equivalence class.

>>>>

Often we take $V$ to be $[n]$, the cyclically ordered set of integers, $1 \ldots n$. In this case, we write $W = (\cP, n)$. We introduce some notation to speak of vertices supporting a propagator, and the set of propagators supported on a vertex set.

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, n)$.
\begin{enumerate}
\item For $p \in \cP$, let $V(p) = \{i_p, i_p+1, j_p, j_p+1\}$ be the set of vertices supporting $p$. Then, for $P \subseteq \cP$, the set $V(P) = \cup_{p \in P} V(p)$ is the vertex support of $P$.
\item For $V \subseteq [n]$, write $\Prop(V) = \{ p \in \cP | V(p) \cap V \neq \emptyset \} $.
\item For $P \subseteq \cP$, define $F(P) = V(P^c)^c$ to be the set of vertices in $[n]$ that do not support any propagators outside the set $P$.
\item The set of vertices that are not in the support of any propagotors is denoted  $F(\emptyset)$. Vertices in this set are called non-supporting.  
\end{enumerate}
\end{dfn}

@@

The first conditions states that there are at least four more vertices than propagators in an admissible Wilson loop diagram. The second imposes an upper bound on how densely the propagators can be fitted in the diagram. The third ensures that ensures that no propagators cross in the interior of the diagram. \note{fixed? earlier, we had $i + 1 < j$ in the definition. When that was fixed, I didn't propagate forward. Or am I missing something else?}In other words, a Wilson loop diagram, $(\cP, n)$ is admissible if and only if $n \geq \cP +4$, and has no crossing propagators, no propagators of the form $p = (i, i+1)$, and no pairs of propagators that start and end on the same pair of non-adjacent edges.

@@

\begin{dfn} \label{subdiagramdfn}
Let $W = (\cP, n)$ be an admissible Wilson loop diagram. The weakly admissible diagram, $W'$ is a subdiagram of $W$, written $W' \subseteq W$, if \bas W' = (P, V); \quad P \subseteq \cP ; \quad V(P) \subseteq V \subseteq [n]\;.\eas
\end{dfn}

@@

\begin{dfn}\label{equivdfn} 
There is an equivalence relationship on the set of admissible Wilson loops diagrams given by the transititve closure the following binary relation: $W = (\cP, n) \sim W'= (\cP', n)$ if
\begin{enumerate}
\item There exist two different exact subdiagrams, $(P, V(P))$ and $(P', V(P'))$ of $W$ and $W'$ respectively such that $V(P) =  V(P')$.
\item The complementary subdiagrams are identical: $(\cP \setminus P, V(P)^c) = (\cP' \setminus P', V(P')^c)$.
\end{enumerate}
\end{dfn}

@@

Each Wilson loop diagram, $W = (\cP, n)$ with $|\cP| = k$ is associated to a $k \times n$ matrix with non-zero real variable entries, called $C(W)$:

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any admissible Wilson loop diagram $W =(\cP, n)$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ such that no subset supports fewer propagators than the vertices it contains.

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from equation \eqref{C(W) dfn}that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subset [n]$,  then one can write $C(W)|_V$ is the restriction of $C(W)$ to the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are real variables, the matrix $C(W)|_V$ does not have full rank if and only if there are fewer rows with non-zero entries than $|V|$. This is equivalent to saying that $V$ is a set of dependent columns of $C(W)$ if and only if $|\Prop(V)| < |V|$. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$. \sanote{Is this suffient for an explanation?}

@@

\begin{dfn}\label{WLDtriangulationdfn}
  Let $W = (\cP, n)$ be an admissible Wilson loop diagram.  The \emph{polygon dissection} associated to $W$, denoted $\tau(W)$, is defined as follows.
  \begin{itemize}
  \item The vertices of $\tau(W)$ correspond to the edges of $W$.
  \item Labeling the vertices of $\tau(W)$ with the edge number of $W$gives a cyclic order to the vertices. Connecting consecutive vertices gives a graph theoretic cycle called the polygon of $\tau(W)$.
  \item Each propagator of $W$ defines a chord edge of $\tau(W)$; specifically,  a propagator $(i,j) \in \cP$ defines a chord connecting the vertices $i$ and $j$ in $\tau(W)$.
  \end{itemize}
\end{dfn}

@@

\begin{lem}\label{tausimpleplanarlem}
If $W = (\cP, n)$ is an admissible Wilson loop diagram, then $\tau(W)$ is a simple planar graph whose outer face is a cycle. It is embedded such that the vertices all lie on this infinite face\footnote{That is, it is an \emph{outerplanar} graph}. These vertices are cyclically ordered, with a choice of first vertex giving it an additional compatible linear order.
\end{lem}

@@

\begin{proof}
We begin by giving an algorithm for the decomposition, then prove its uniqueness. Let $W = (\cP, n)$, with $|\cP| = k$.

@@

Every triangulated piece $t$ of an admissible Wilson loop diagram $W$ corresponds to a subdiagram of $W$ by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $W_P = (P, V(P))$.  Conversely, given a subdiagram $W_P = (P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, as follows:
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ defined by cyclically consecutive elements of $V(P)$ as a subset of $[n]$
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$
  \item along with the outer edges of $\tau(W)$ for which both their end points are in the vertex set
  \end{itemize}
\end{itemize}

\begin{eg}
For example, consider the following Wilson loop diagram, $W$, and its corresponding chord diagram, $\tau(W)$. 
\bas W = \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawWLD{10}{1.5}
	\drawnumbers
	\drawprop{1}{0}{4}{0}
	\drawprop{2}{0}{4}{-1}
	\drawprop{4}{1}{8}{0}
	\drawprop{8}{1}{1}{-1}
	\drawprop{5}{0}{7}{0}
		\end{scope}
	\end{tikzpicture}  \; \quad
\tau(W) = \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawpolypart{10}{1.5}
    \drawnumbersshift
    \drawchord{1}{4}
    \drawchord{2}{4}
    \drawchord{4}{8}
    \drawchord{8}{1}
    \drawchord{5}{7}
	\end{scope}
	\end{tikzpicture} \eas

The set of propagators $P= \{(1,4), (4, 8), (1,8) \}$, define an exact subdiagram of $W$. The vertex set of $t$ are the vertices labeled $\{1, 2,3, 4, 5, 8\}$. The subgraph $t$ is then

\bas t = 
\begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
\pgfmathsetmacro{\n}{10}
\pgfmathsetmacro{\radius}{1.5}
\pgfmathsetmacro{\angle}{360/\n}
    \foreach \i in {1,2,3,4} {
      \draw (\angle*\i+ \angle/2:\radius) node {$\bullet$}; }
    \foreach \i in {2,3,4} {
     \pgfmathsetmacro{\x}{\angle*\i - \angle/2}
      \pgfmathsetmacro{\concave}{((\n-1.5)/\n)}
      \draw (\x:\radius cm) .. controls (\angle *\i: \concave* \radius cm) .. (\x + \angle:\radius cm);
      \draw (\angle*8+ \angle/2:\radius) node {$\bullet$};
}    
\foreach \i in {1,2,3,4,8} {
  \pgfmathsetmacro{\x}{\angle*\i + \angle/2}
  \draw (\x:\radius*1.15) node {\footnotesize \i};
}
    \drawchord{1}{4}
    \drawchord{2}{4}
    \drawchord{4}{8}
    \drawchord{8}{1}
	\end{scope}
	\end{tikzpicture} \eas

Note that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in $W_P$ itself.  In particular $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$. 
\end{eg}

@@

Conversely, suppose we have an exact subdiagram $(P, V(P))$ of $W = (\cP, n)$ supported on $|V(P)| = |P|+3$ vertices, and let $t$ be the subgraph of $\tau(W)$ corresponding to $(P,V(P))$.

@@

\begin{cor} \label{uniqueproppartitioncor}
Any admissible Wilson loop diagram $W = (\cP, n)$ can be uniquely decomposed into maximal exact subdiagrams. These maximal subdiagrams partition $\cP$.
\end{cor}

@@

\begin{dfn}\label{def prop flat} 
Let $W = (\cP,n)$ be an admissible Wilson loop diagram and $P \subseteq \cP$. The set $F(P) := V(P^c)^c$ is called the {\em propagator flat} of $P$.
\end{dfn}
Thus in Lemma \ref{contractsubdiaglem} above we would contract $M(W)$ by the propagator flat $F(P^c)$ of the {\em complement} of $P$ in order to study $(P,V(P))$. The justification for this notation is given by (1) of the next lemma: the set $F(P)$ consists of vertices which {\em only} support propagators in $P$ (or no propagators at all).

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, n)$ be a Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $F(P^c)$ is not an independent flat.
\item If $(P,V(P))$ is a maximal exact subdiagram in $W$, then $F(P^c)$ is a cyclic flat.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                          \item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate}
\end{lem}

@@

\begin{lem} \label{contractsubdiaglem} \cite[Theorem 3.33]{wilsonloop} 
Let $W = (\cP, n)$ be an admissible Wilson loop diagram and $P \subseteq \cP$. If the set $V(P)^c$ has rank $|P^c|$, then the matroid defined by the subdiagram $(P, V(P))$ is equal to the contraction $M(W)/V(P)^c$.
\end{lem}

@@

\begin{thm} \label{exactuniformthm}
Let $W':= (P, V(P))$ be a subdiagram of an admissible Wilson loop diagram $W= (\cP, n)$. Then $W'$ is an exact subdiagram if and only if $M(W')$ is a uniform matroid of rank $|P|$.
\end{thm}

@@

\begin{dfn}
Given a Wilson loop diagram $W = (\cP, n)$, define the positroid cell associated to a Wilson loop diagram, $\Sigma(W)$, to be the cell in the CW complex on $\Gr(|\cP|, n)$ defined by the positroid $M(W)$.
\end{dfn}

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, n)$ and $W'= (\cP', n)$ be two Wilson Loop diagrams. They define the same matroid if and only if $W \sim W'$.
\end{thm}

@@

Assume that $W$ and $W'$ are equivalent. Without loss of generality, write $W = (P \cup R, n)$ and $W' = (P \cup R', n)$, where $P \subset \cP \cap \cP'$ and $(R, V(R))$ and $(R', V(R'))$ are two maximally exact subdiagrams, with $R \neq R'$, but $V(R) = V(R')$. If this is not the case, one may always find a family of diagram, $\{W_i\}$ satisfying this condition and forming a transitive chain connecting $W$ to $W'$ in the equivalence class.
